---
title: "Starter Analysis Exercise"
---

Note: This document contains code needed for each part of the Starter Analysis exercise for the MADA Spring 2026 course. Each part is listed under its corresponding part name. As a disclaimer, AI was used for code brevity and assistance in fixing errors in code. `Cmd + option + I` was used to create chunks for code sections (mac option).

## Part 1

### Package loading and install

```{r}
# install dslabs package (run once in the console if not already installed)
# install.packages("dslabs")

# load required packages
# tidyverse added for later steps with data processing
library("dslabs")
library("tidyverse")
```

## Show dataset structure

```{r}
# only run the next command interactively, not in a script
# help(gapminder)

# get an overview of data structure
str(gapminder)
```

```{r}
# get a summary of data
summary(gapminder)
```

```{r}
# determine the type of object gapminder is
class(gapminder)
```

### Data processing

### Process Africa subset

```{r}
# subset gapminder data to rows where continent is "Africa"
africadata <- gapminder[gapminder$continent == "Africa", ]

# inspect structure (should show 2907 observations; troubleshoot if not)
str(africadata)

# check summary statistics 
summary(africadata)
```

```{r}
# keep only infant_mortality and life_expectancy
africa_infant_mortality <- africadata[, c("infant_mortality", "life_expectancy")]

# keep only population and life_expectancy
africa_population <- africadata[, c("population", "life_expectancy")]

# inspect structure + summaries
str(africa_infant_mortality)
summary(africa_infant_mortality)

str(africa_population)
summary(africa_population)
```

### How to plot the data

# 

```{r}
# life expectancy vs infant mortality (points)
plot(
  x = africa_infant_mortality$infant_mortality,
  y = africa_infant_mortality$life_expectancy,
  xlab = "Infant Mortality",
  ylab = "Life Expectancy",
  main = "Infant Mortality vs Life Expectancy (Africa)"
)
```

```{r}
# life expectancy vs population (points) with log-scaled x-axis
plot(
  x = africa_population$population,
  y = africa_population$life_expectancy,
  xlab = "Population (log scale)",
  ylab = "Life Expectancy",
  main = "Population vs Life Expectancy (Africa)",
  log = "x"
)
```

The `infant mortality plot` shows a negative correlation with life expectancy, while the population plot shows a general positive relationship. The plots appear to in a streaking form because individual countries are counted across multiple points. In the `population vs life expectancy (Africa) plot`, we can see that for individual country populations, as population increases, so does life expectancy.

```{r}
# count missing infant_mortality values for each year
tapply(africadata$infant_mortality, africadata$year, function(x) sum(is.na(x)))
# subset to year 2000
africa_2000 <- africadata[africadata$year == 2000, ]
str(africa_2000)
summary(africa_2000)
```

```{r}
fit1 <- lm(life_expectancy ~ infant_mortality, data = africa_2000)
summary(fit1)

fit2 <- lm(life_expectancy ~ population, data = africa_2000)
summary(fit2)
```

`Fit1` shows that `infant mortality` shows a statistically significant negative relationship with `life expectancy` **(p = 2.83\^-08)**, explaining approximately **47% of the variation in life expectancy** among African countries in 2000. In contrast, `population size` shows no significant association with life expectancy (p = 0.616). The model fit is also highly significant with a prediction error rate of 6.221 years (F= 43.48, p= 2.825\^-08).

`Fit2` indicates that there is no statistically significant relationship between `population size` and `life expectancy` (estimate =2.756\^-08; p=0.616). The model is not significant and does not explain the variation in life expectancy (R2= 0.005176; adjusted value of -0.01513; F= 0.2549, **p= 0.6159**). Predictor error is large for this dataset at 8.524 years.
